Quantifying Stability and Change in Personal Culture Using Panel Data

### 2023-09-20 manuscript.Rmd
small changes in wording

@@

The second model summarized by Kiley and Vaisey, an “Active Updating Model” (AUM), posits that people continually update their personal culture as they move through life. This model suggests people make new cultural meanings, and therefore durably change their personal culture, as they encounter new social environments, discourses, and information [@gross2009; @swidler2001]. This model underlies models of cultural diffusion [@christakis2010], attitude alignment [@dellaposta2020], and polarization [@bail2018], among others. It is also implicit in most studies of whether specific experiences, changes in social roles, or political events, affect personal culture [@gelman2021; @visser2004; @slothuus2021].

@@

In contrast to the tournament of models approach, quantifying change this way allows us to explore variation in the relative importance of intrapersonal change across questions that all both show evidence of change. For example, Kiley and Vaisey [-@kileyvaisey2020] found that confidence in the press and confidence in religious leaders were both characterized by active updating, our results show that intrapersonal change is much more important for explaining variance in confidence in the press (0.164) than confidence in religion (0.049).

@@

This work has tended to focus on the fact that college-educated Americans give responses that are less likely to be affected by measurement error or short-term influences [@zaller1992; @alwin2003]. However, this focus on the difference in the non-systematic (residual) component of variance across groups obscures the fact that the relative importance of different components of systematic variance might differ across these two groups as well.

It could be the case that education, chronic knowledge, and attention to the political sphere makes intrapersonal change more important in explaining differences among people with higher levels of education. For example, these people might be more likely to make systematic changes in response to the emergence of new information, new issues, or political realignments, while those without chronic knowledge display more variance around their baselines without their baselines changing much [@zaller1992]. In contrast, it could be the case that the stability demonstrated by educated respondents reflects the fact that they have already formed durable opinions and are relatively closed off to new information. College could serve as a formative experience that solidifies attitudes for people who attend. But among people who do not attend college, later life experiences might prove more important in forming or changing attitudes, as these experiences potentially provide information that college-educated peers received earlier.

@@

[^wave]: A small number of respondents report different highest degrees at wave 1, wave 2, and wave 3. Some of this is due to measurement error, and some of it is due to a small number of people obtaining a higher degree during the four years of the panel. Estimating the panel with highest at wave 3 or highest degree reported across the panel produces functionally identical results.

@@

Figure XXX plots the distribution of differences between people with less than a bachelor's degree and people with at least a bachelor's degree at wave 1 of the panel. Values greater than 0 would indicate that intrapersonal change is more important in explaining the responses of college graduates than those of people without a college degree, while values less than 0 would indicate that intrapersonal change is less important for explaining variation among people without a college degree than those with a college degree.

@@

This paper set out to clarify previous findings about the prevalence of durable change in personal culture -- the attitudes, beliefs, values, behaviors, and preferences that manifest at the individual level. Relying on an approach that quantified the amount of systematic variance attributable to either intrapersonal change or interpersonal differences at baseline, we found a consistent pattern across data sets previously used in this debate. We see evidence of intrapersonal change over time on almost all questions, with some measures of personal culture displaying quite high levels of intrapersonal change relative to the amount of variance between people. On some questions, such as the factors that affect one’s life satisfaction or attitudes about government benefits, it is highly plausible that differences in peoples’ experiences as adults are the primary explanation for differences in personal culture.

However, the amount of intrapersonal change is often small relative to interpersonal differences (less than 10 percent on average across all surveys). On many questions, the amount of systematic variance attributable to intrapersonal change is essentially zero. These include questions about civil liberties, abortion, inter-individual (kk note: generalized?) trust, or sense of civic duty. On these questions and the large number of others with a low proportion of systematic variance attributable to intrapersonal change, there does not appear to be sufficient cultural change during adulthood for experiences and transitions between social locations to be the primary explanation for why adults end up with different personal cultures. For these items, the explanation for personal cultural differences seems to reside principally in early-life experiences.

@@

To be clear, none of our results should be interpreted as suggesting that adults do not change or are incapable of changing. On all questions, some members of the sample change their personal culture at least a little bit. Our results also should not be taken to imply that intrapersonal change is not important or consequential. Small numbers of changers can have important consequences in the aggregate. A shift from 49 percent supporting gay marriage to 51 percent supporting it can have significant political ramifications, even as most respondents remain fixed in their preferences. And our results should not detract from work trying to estimate the mechanisms through which personal cultural differences --however small-- emerge among adults. Just because change is small does not mean the mechanisms that produce is are not important to understand.

@@

Furthermore, the results we present related to education and political views suggest that the relative importance of different theoretical processes might differ by group. In line with life course perspectives, the results presented here suggest that importance of different experiences for attitude change are potentially contingent on previous experiences. For example, to the extent that we want to explain why people have different personal culture, the answer is likely different for people who have a college degree and those who do not. Mid-life experiences appear to have more significant implications for the personal culture for people without a college degree than for people with a college degree.

@@

However, because our results suggest that the role of formative experiences compared to contemporaneous environments differs by a person's level of education, changes in educational attainment over time, as well as changes in the social structuring of the life course might be shifting the relative importance of these two factors over time. If we observe further differences across socio-demographic groups like education, social class, and race/ethnicity in the relative importance of these processes, this would argue for a greater incorporation the interaction of demographic processes with processes of individual cultural change into our explanations of aggregate cultural change.

@@

A major takeaway from the meta-analysis is that greater response resolution –-the number of response options survey respondents are given-– is associated with a greater share of variance being explained by within-individual change. Again, it is possible that this reflects different kinds of issues being measured with different kinds of scales. But it could also suggest that within-person change, when it happens, is relatively small –-moving from “agree” to “strongly agree”, rather than “strongly agree” to “strongly disagree,” for example-– and therefore easier to detect with greater resolution.

@@

Similarly, in assuming a linear trajectory for change, our approach assumes that durable change is unidirectional. To the extent that people make durable change as a result of a social transition and then return to a previous position later in life as a result of a second transition (marriage and divorce, for example), this is captured as residual within-person variance, rather than one of our two broad theoretical processes. This is a potentially reasonable assumption on short panels -- change that does not last two years is hard to classify as durable -- but it might not be true in longer panels. Other researchers with more flexible definitions of "durable change" might potentially identify more durable change even in the same population.

Finally, while we examine an incredible broad array of measures of personal culture, from assessments about the relative importance of different features when buying a new car and policy preferences to religious beliefs and core values, we are limited by the kinds of questions that are asked in panel surveys and the contexts in which these surveys were administered (Australia, Germany, Switzerland, the United States, and the United Kingdom). This means our findings tend to reflect questions about general (national) politics, gender roles, immigration and race relations, and general well being. Our data does not cover some dimensions of culture that might be relevant for other aspects of sociological exploration (artistic tastes, leisure activities, and time use, for example).

# Implications for Future Research

@@

While we focused on education as a determinant of the explanatory power of intrapersonal change, our approach can readily be extended to other socio-demographic categories that might affect intrapersonal change -- race/ethnicity, religiosity, social class, gender, and political affiliation, for example -- or their intersections. Researchers with different data might be able to evaluate the relative importance of intrapersonal change and interpersonal differences across contexts as well, either national contexts if the same questions are administered across countries or organizational contexts if data traces students within schools, workers within workplaces, or congregation members within congregations.

>>>>

## Change and Stability in Personal Culture

@@

The second model summarized by Kiley and Vaisey [-@kiley2020], an “Active Updating Model” (AUM), posits that people continually update their personal culture as they move through life. This model suggests people make new cultural meanings, and therefore durably change their personal culture, as they encounter new social environments, discourses, and information [@gross2009; @swidler2001]. This model underlies models of cultural diffusion [@christakis2010], attitude alignment [@dellaposta2020], and polarization [@bail2018], among others. It is also implicit in most studies of whether specific experiences, changes in social roles, or political events, affect personal culture [@gelman2021; @visser2004; @slothuus2021].

@@

In contrast to the tournament of models approach, quantifying change this way allows us to explore variation in the relative importance of intrapersonal change across questions that all both show evidence of change. For example, Kiley and Vaisey [-@kiley2020] found that confidence in the press and confidence in religious leaders were both characterized by active updating, our results show that intrapersonal change is much more important for explaining variance in confidence in the press (0.164) than confidence in religion (0.049).

@@

This work has tended to focus on the fact that college-educated Americans give responses that are less likely to be affected by measurement error or short-term influences [@zaller1992; @alwin2007]. However, this focus on the difference in the non-systematic (residual) component of variance across groups obscures the fact that the relative importance of different components of systematic variance might differ across these two groups as well.

It could be the case that education, chronic knowledge, and attention to the political sphere makes intrapersonal change more important in explaining differences among people with higher levels of education. For example, these people might be more likely to make systematic changes in response to the emergence of new information, new issues, or political realignments, while those without chronic knowledge display more variance around their baselines without their baselines changing much [@zaller1992]. In contrast, it could be the case that the stability demonstrated by educated respondents reflects the fact that they have already formed durable opinions and are relatively closed off to new information. College could serve as a formative experience that solidifies personal culture for people who attend. But among people who do not attend college, later life experiences might prove more important in forming or changing personal culture, as these experiences potentially provide information that college-educated peers received earlier.

@@

[^wave]: A small number of respondents report different highest degrees at wave 1, wave 2, and wave 3. Some of this is due to measurement error, and some of it is due to a small number of people obtaining a higher degree during the four years of the panel. Estimating the panel with highest at wave 1 or highest degree reported across the panel produces functionally identical results.

@@

Figure XXX plots the distribution of differences between people with less than a bachelor's degree and people with at least a bachelor's degree at wave 1 of the panel. Values greater than 0 would indicate that intrapersonal change is more important in explaining the responses of college graduates than those of people without a college degree, while values less than 0 would indicate that intrapersonal change is less important for explaining variation among people with a college degree than those without a college degree.

@@

This paper set out to clarify previous findings about the prevalence of durable change in personal culture -- the attitudes, beliefs, values, preferences, and behaviors that manifest at the individual level. Relying on an approach that quantified the amount of systematic variance attributable to either intrapersonal change or interpersonal differences at baseline, we found a consistent pattern across data sets previously used in this debate. We see evidence of intrapersonal change over time on almost all questions, with some measures of personal culture displaying quite high levels of intrapersonal change relative to the amount of variance between people. On some questions, such as the factors that affect one’s life satisfaction or attitudes about government benefits, it is highly plausible that differences in peoples’ experiences as adults are the primary explanation for differences in personal culture.

However, the amount of intrapersonal change is often small relative to interpersonal differences (less than 10 percent on average across all surveys). On many questions, the amount of systematic variance attributable to intrapersonal change is essentially zero. These include questions about civil liberties, abortion, generalized trust, or sense of civic duty. On these questions and the large number of others with a low proportion of systematic variance attributable to intrapersonal change, there does not appear to be sufficient cultural change during adulthood for experiences and transitions between social locations to be the primary explanation for why adults end up with different personal cultures. For these items, the explanation for personal cultural differences seems to reside principally in early-life experiences.

@@

To be clear, none of our results should be interpreted as suggesting that adults do not change or are incapable of changing. On all questions, some members of the sample change their personal culture at least a little bit. Our results also should not be taken to imply that intrapersonal change is not important or consequential. Small numbers of changers can have important consequences in the aggregate. A shift from 49 percent supporting gay marriage to 51 percent supporting it can have significant political ramifications, even as most respondents remain fixed in their preferences. And our results should not detract from work trying to estimate the mechanisms through which personal cultural differences --however small-- emerge among adults. Just because change is small does not mean the mechanisms that produce it are not important to understand.

@@

Furthermore, the results we present related to education and political views suggest that the relative importance of different theoretical processes might differ by group. In line with life course perspectives, the results presented here suggest that importance of different experiences for cultural change are potentially contingent on previous experiences. For example, to the extent that we want to explain why people have different personal culture, the answer is likely different for people who have a college degree and those who do not. Mid-life experiences appear to have more significant implications for the personal culture for people without a college degree than for people with a college degree.

@@

However, because our results suggest that the role of formative experiences compared to contemporaneous environments differs by a person's level of education, changes in educational attainment over time, as well as changes in the social structuring of the life course might be shifting the relative importance of these two factors over time. If we observe further differences across socio-demographic groups like education, social class, and race/ethnicity in the relative importance of these processes, this would argue for a greater incorporation of the interaction of demographic processes with processes of individual cultural change into our explanations of aggregate cultural change.

@@

A major takeaway from the meta-analysis is that greater response resolution –-the number of response options survey respondents are given-– is associated with a greater share of variance being explained by within-individual change. Again, it is possible that this reflects different kinds of issues being measured with different kinds of scales. But it could also suggest that within-person change, when it happens, is relatively small –-moving from “agree” to “strongly agree,” rather than “strongly agree” to “strongly disagree,” for example-– and therefore easier to detect with greater resolution.

@@

Similarly, in assuming a linear trajectory for change, our approach assumes that durable change is unidirectional. To the extent that people make durable change as a result of a social transition and then return to a previous position later in life as a result of a second transition (marriage and divorce, for example), this is captured as residual within-person variance, rather than one of our two broad theoretical processes. This is a potentially reasonable assumption on short panels --change that does not last two years is hard to classify as durable-- but it might not be true in longer panels. Other researchers with more flexible definitions of "durable change" might potentially identify more durable change even in the same population.

Finally, while we examine an incredible broad array of measures of personal culture, from assessments about the relative importance of different features when buying a new car and policy preferences to religious beliefs and core values, we are limited by the kinds of questions that are asked in panel surveys and the contexts in which these surveys were administered (Australia, Germany, Switzerland, the United Kingdom and the United States). This means our findings tend to reflect questions about general (national) politics, gender roles, immigration and race relations, and general well being. Our data does not cover some dimensions of culture that might be relevant for other aspects of sociological exploration (artistic tastes, leisure activities, and time use, for example).

## Implications for Future Research

@@

While we focused on education as a determinant of the explanatory power of intrapersonal change, our approach can readily be extended to other socio-demographic categories that might affect intrapersonal change --race/ethnicity, religiosity, social class, gender, and political affiliation, for example-- or their intersections. Researchers with different data might be able to evaluate the relative importance of intrapersonal change and interpersonal differences across contexts as well, either national contexts if the same questions are administered across countries, or organizational contexts if data traces students within schools, workers within workplaces, or congregation members within congregations.
